%%
%% patent_analysis_tool.tex — Programmatic Patent Document Analysis Tool
%% CS 294-184 Final Project Part 2
%% Based on sample-sigconf-authordraft.tex from the acmart package.
%%
\documentclass[sigconf]{acmart}
\usepackage{microtype}
\usepackage{listings}
\usepackage{xcolor}
\emergencystretch=1em

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.
\setcopyright{acmlicensed}
\copyrightyear{2026}
\acmYear{2026}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2018/06}

%% Listings style for DSL code examples
\lstdefinelanguage{PatentDSL}{
  morekeywords={AND, OR, NOT, section, contains, cpc, meta, paragraph},
  sensitive=false,
  morestring=[b]",
}
\lstset{
  language=PatentDSL,
  basicstyle=\ttfamily\small,
  keywordstyle=\bfseries,
  stringstyle=\color{gray},
  frame=single,
  breaklines=true,
  captionpos=b,
}

%%
%% Start of document
%%
\begin{document}

\title{A Programmatic Document Analysis Environment for Patent Examination}

\author{Shanaya Malik}
\email{shanaya_malik@berkeley.edu}
\affiliation{%
  \institution{University of California, Berkeley}
  \city{Berkeley}
  \state{California}
  \country{USA}}

\author{Kelvin Zhao}
\email{kelvinyz@berkeley.edu}
\affiliation{%
  \institution{University of California, Berkeley}
  \city{Berkeley}
  \state{California}
  \country{USA}}

\renewcommand{\shortauthors}{Malik \& Zhao}

%% ====================================================================
%%  ABSTRACT
%% ====================================================================
\begin{abstract}
% TODO
\end{abstract}

\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10003120.10003121.10003122.10010853</concept_id>
  <concept_desc>Human-centered computing~Empirical studies in HCI</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10003120.10003121.10003124.10010870</concept_id>
  <concept_desc>Human-centered computing~HCI design and evaluation methods</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
  <concept_desc>Software and its engineering~Functional languages</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Human-centered computing~Empirical studies in HCI}
\ccsdesc[300]{Human-centered computing~HCI design and evaluation methods}
\ccsdesc[300]{Software and its engineering~Domain-specific languages}

\keywords{patent examination, prior art search, domain-specific language,
document analysis, contextual inquiry, sensemaking, information work, HCI}

\maketitle


%% ====================================================================
%%  SECTION 1 — MOTIVATION  (1–2 pages)
%% ====================================================================
\section{Motivation}

\smallskip\noindent\textbf{The stakes of patent examination.}
Patent examination is a high-stakes, knowledge-intensive activity in which
trained examiners must locate prior art---existing patents, publications, or
other disclosures---that \emph{anticipates} or \emph{obviates} each claim in a
pending patent application. Because issued patents carry legal force, the
quality of examination directly affects the validity of the intellectual
property record and the downstream costs of litigation and re-examination.
Yet the core task of finding prior art remains largely manual: an examiner
reads the claims, identifies the key technical concepts, formulates search
queries, retrieves candidate documents, reads those documents to find relevant
passages, and then records exactly \emph{where} in each document the evidence
appears so it can be cited in an office action. For complex technical art
units---software, biomedical devices, quantum computing---this cycle can span
dozens of documents and hundreds of individual claim elements per application.

\smallskip\noindent\textbf{Why existing tools are insufficient.}
Current search infrastructure, including the USPTO's internal EAST and WEST
systems as well as commercial tools such as PatSnap and Derwent Innovation,
treats this workflow at the document level. A query returns a ranked list of
patent documents; the examiner then opens each document and manually locates
the relevant section. Tools surface keyword hit \emph{counts}, but do not
distinguish whether a hit appears in the abstract, the background, the
detailed specification, or the claims---sections that carry very different
evidentiary weight. Passage-level retrieval, section-aware filtering, and
metadata-constrained queries are not available as first-class capabilities in
any tool accessible to working examiners.

\smallskip\noindent\textbf{Evidence from practitioners.}
Through a contextual inquiry study with ten practicing patent examiners (reported
in full in Sections~3--4), we identified five recurring needs that current
tooling does not address. Two of these needs are particularly relevant to the
design of a document analysis environment. First, examiners need
\emph{section-aware retrieval}: P05 explicitly described wanting hits in the
background and detailed description sections, not in the abstract or claims,
because the specification is where the technical teaching that maps to claim
elements is found. Section type is a structural property of every patent
document, yet no current tool exposes it as a query parameter. Second,
examiners need \emph{pinpoint citation}: once a relevant passage is located,
examiners must record its position in the document for citation in the office
action. For post-2001 USPTO publications, this position is a numbered
paragraph anchor (e.g., \texttt{[0042]}); for pre-AIA granted patents, it is a
column-and-line number. This pinpoint address is embedded in the document's
structure, yet retrieving by it is not possible in any current tool.

\smallskip\noindent\textbf{The root cause: documents as opaque blobs.}
A further structural problem unifies all five needs: patent documents are
treated as opaque text blobs. Queries operate over full-text indexes; results
are document-level rankings. The rich internal structure of a patent---its
sections, numbered paragraphs, claim hierarchy, and embedded
metadata---is invisible to the query layer. An examiner who wants to find all
passages in the detailed specification of a particular assignee's patents that
discuss a specific technical mechanism must open each document, scroll to the
relevant section, and scan manually. This is not a failure of individual tools
but a consequence of the retrieval model: when documents are the unit of
retrieval, there is no path to passage-level precision.

\smallskip\noindent\textbf{Our approach and expected outcomes.}
We propose a different approach. Rather than building another document-level
search interface, we expose patent documents as \emph{structured, queryable
data}. Our system parses each patent into a \texttt{Document~$\to$~Section~$\to$~Passage}
hierarchy and extracts all available metadata fields. Examiners then write
queries in a minimal domain-specific language (DSL) that combines structural
filters---\texttt{section:SPECIFICATION}, \texttt{paragraph:0042}---with
content filters and metadata constraints, all composable with Boolean
operators. The query executes over the passage-level representation and returns
matched passages with their full provenance: which document, which section,
which paragraph anchor, and neighboring context.

\smallskip\noindent\textbf{Why a DSL fits this population.}
This design is motivated by two key insights from our need-finding. First,
patent examiners are not casual users: they are technical specialists who
write Boolean search strings as a core job skill, and several participants
described current query interfaces as too limited rather than too complex
(P04, P07, P10). A textual DSL is therefore appropriate for this population
in a way it might not be for a general consumer audience. Second, the
iterative, refinement-based search strategy that examiners actually use (Need~5)
is precisely what a composable query language supports: an examiner can start
with a broad \texttt{section:CLAIMS} filter, observe the results, narrow with
\texttt{AND contains:"neural interface"}, and further restrict with
\texttt{AND NOT paragraph:0001} to skip boilerplate preambles. Each refinement
is a surgical edit to the query string, not a round-trip through a separate
GUI tool.

The contributions of this paper are as follows:
\begin{itemize}
  \item We describe a passage-level document model for patent text that
    exposes section type, numbered paragraph anchors, claim structure, and
    rich patent metadata as first-class query targets, grounded in the
    structural needs identified through contextual inquiry.
  \item We define a minimal DSL for composable patent queries, supporting
    \texttt{section:}, \texttt{contains:}, \texttt{paragraph:},
    \texttt{cpc:}, and \texttt{meta.} filters with full Boolean logic
    (\texttt{AND}, \texttt{OR}, \texttt{NOT}, parentheses), and describe its
    design rationale in terms of the five needs from our study.
  \item We present a prototype implementation---a FastAPI backend with a
    React frontend---that parses USPTO-format patent documents (both
    plain-text and text-extractable PDF) into the passage-level model and
    executes DSL queries with passage-level results, including neighboring
    context and match provenance.
  \item We characterize the open technical challenges in this design space,
    including document parsing fidelity, query expressiveness limits, and
    the performance tradeoffs of in-memory passage indexing at corpus scale.
\end{itemize}


%% ====================================================================
%%  SECTION 2 — RELATED WORK  (1.5–2 pages)
%% ====================================================================
\section{Related Work}

\subsection{Sensemaking and Information Foraging in Professional Technical Work}

% TODO

\subsection{Tools for Patent Search and Document Analysis}

% TODO

\subsection{Domain-Specific Languages for Structured Document Access}

% TODO

\subsection{Contextual Inquiry for Studying Specialized Workflows}

% TODO


%% ====================================================================
%%  SECTIONS 3–4 — NEED FINDING  (from Part 1)
%% ====================================================================
\section{User Study Design and Procedure}

The study was designed to address the research question: \emph{What kinds of
problems do technical patent examiners face during the iterative process of
discovering, filtering, and synthesizing prior art to evaluate the novelty of
complex technical claims?} This question was scoped to examiners in technical
art units (such as software, biomedical engineering, or quantum computing),
where the demands of translating between legal claim language and technical
implementation are particularly demanding. We chose not to pre-specify
hypotheses, since the study was exploratory, need-finding.

The participants were recruited through purposive and snowball sampling,
beginning with two professional contacts who shared a recruitment message
with colleagues and provided us with additional contacts. To supplement the
remaining number of participants, we used cold messaging on LinkedIn to target
individuals whose profiles indicated professional experience in technical
patent examination. More specifically, the inclusion criteria required
participants to be at least 18 years old, fluent in English, and currently or
formerly employed in patent examination, with experience evaluating complex
technical claims. We excluded individuals who exclusively examine
non-technical patents and anyone unable to demonstrate their workflow using
only publicly available documents.

Each session lasted approximately 45 minutes and was conducted remotely via
Zoom. The sessions began with a brief unrecorded introduction of the
participant and researcher, followed by an overview of the project scope and
agenda for the session. The core of the session (${\sim}$30--45 minutes,
recorded) was a contextual inquiry following the Master-Apprentice model, in
which the participant shared their screen and demonstrated their prior-art
search and synthesis process for a given public patent while thinking aloud,
with clarifying questions asked at natural pause points~\cite{Beyer1997}.
This was followed by a synthesis segment focused on how the participant maps
and documents the relationship between prior art and specific claim elements,
and concluded with a closing discussion addressing the most time-consuming and
mentally demanding aspects of the workflow.

Each session produced three data artifacts stored on a password-protected
Berkeley bDrive account: (1)~a Zoom screen-and-audio recording capturing
screen activity, search queries, and think-aloud narration; (2)~typed session
notes taken in real time; and (3)~a post-session debrief document recording
emerging patterns. The recordings were transcribed within Zoom itself and
reviewed for accuracy. The participants were assigned coded identifiers (P01,
P02, etc.), and we consolidated each participant's transcript, session notes,
and debrief into a single document as part of our input to our reflexive
thematic analysis~\cite{Braun2006}.


\section{User Study Results}

\subsection{Analysis Process}

We analyzed the consolidated session documents using reflexive thematic
analysis following Braun and Clarke~\cite{Braun2006}. Each researcher
independently conducted an initial round of open coding on their assigned
participant's contextual inquiries, applying purely descriptive labels to
segments that capture actions, reasoning, breakdowns, and workarounds within
each patent examiner's workflow. Since our data include both think-aloud
narration and question answers, as well as screen-recorded behavior, our
coding explicitly distinguishes between what participants said and what
actions they took, allowing us to capture both direct information from
explanations and implicit information from instances where difficulties were
not verbally articulated. As a result, many codes described domain-specific
interactions, such as refining search queries, tracking claim coverage across
references, and using specific internal tools to manage information across
documents.

After independent coding, we compared our code sets across assigned
participants and iteratively refined them through discussion, focusing first
on resolving differences in interpretation and initial grouping, before
collaboratively identifying higher-level patterns that showed recurring needs
across multiple participants. We treated this stage as an iterative process,
merging overlapping codes and preserving any meaningful differences reflecting
underlying issues in the participants' workflows. We grouped related codes
into candidate themes, tightened their definitions, and verified that each
theme was supported by evidence from multiple participants. The five themes we
present below represent the needs that emerged most consistently across our
participant pool.

\subsection{Findings}

\smallskip\noindent\textbf{Need 1: Examiners lack systematic support for
generating and managing the synonyms needed to translate legal claim language
into effective search queries.}
A central challenge across sessions was the gap between claim terminology and
the terminology used to retrieve relevant prior art; as P04 explained, the
applicant ``comes up with their own term, so we try to come up with terms
that are going to be describing the same thing.'' P06 described this as a
problem of implied language: a term like \emph{modder} in a medical device
patent could return results about computer modification, and a term like
\emph{wearable} might not appear in a relevant document that uses a synonym,
meaning those documents would be filtered out. Examiners relied on a patchwork
of ad hoc resources, including a shared synonym crosswalk tool (P04), Google
searches (P04), terminology from prior specifications, and personal
experience. P06 described a case where an applicant's term for an ophthalmic
structure appeared only in their own application, requiring consultation with
a senior examiner to identify standard terminology. P07 noted that while
examiners are taught about synonyms and Boolean connectors, applying this in
unfamiliar art areas requires substantial independent research, alongside
occasionally having to consult a senior examiner. Similarly, P10 described
getting ``stuck with the synonyms'' when attempting to expand search queries
in unfamiliar or highly technical applications, reinforcing that synonym
generation remains an unsupported and cognitively demanding step. Junior
examiners were particularly disadvantaged, lacking the tacit vocabulary that
senior examiners described as built through years of repetition (P01, P05,
P06).

\smallskip\noindent\textbf{Need 2: Current search tools do not adequately
support filtering by filing date, leading to wasted effort and rework.}
Every participant referenced the effective filing date as a hard constraint on
admissible prior art, yet existing tools provided limited support for
enforcing it during search. P05 described completing an entire office action
based on a reference that turned out to postdate the application's effective
filing date, requiring a complete redo. P06 identified a related problem in
which search systems frequently confuse publication dates with filing dates,
and also confuse non-provisional filing dates with earlier provisional dates
that establish actual priority. P06 called this ``definitely a very good fix
that can be done.'' P01 noted that the rules for effective filing dates differ
between U.S.\ and foreign applications, making even manual checks non-trivial.
P08 described the filing-date constraint as deeply embedded in the workflow,
requiring cross-referencing across domestic, provisional, and foreign filings,
and P03 had initially demonstrated this in action, showing that of seven
initial results, only one had a usable date (but determining this required a
manual, time-consuming inspection of each).

\smallskip\noindent\textbf{Need 3: The transition from search to claim
mapping relies on entirely manual, ad hoc methods with little integrated tool
support.}
After identifying promising prior art, examiners must document how specific
passages address specific claim elements, which is a process called claim
mapping. Every participant performed this using personal, improvised methods
outside the search tool. P01 maintained a Word table; P03 copied claims as
bullet points and mapped inline; P04 used a personal claim chart; and P05
copied claims into a Word document with parentheses after each element and
drew dependency trees by hand. P06 described using pen and paper for claim
trees and explained that the in-house software's mapping feature was so
inaccurate that ``pretty much no one at the USPTO'' trusts it---errors in
claim dependencies can distort the entire analysis. P07 described the process
as lacking elegant tooling, stating that ``it's a very manual process,
[examiners] just start plopping it in there.'' P08 described printing claims
on paper and annotating them physically to manage cognitive load across
simultaneous cases, and P05 explicitly identified this as a design
opportunity, stating that integrating claim mapping into the search tool could
save a significant amount of time and energy for examiners.

\smallskip\noindent\textbf{Need 4: Examiners struggle to locate and interpret
relevant information within retrieved documents.}
Examiners reported that tools display keyword hit counts without
distinguishing whether hits appear in the specification (the detailed
description most useful for mapping), the abstract, or the claims. P05 wanted
hits specifically in the background and description sections, not in the
abstract or claims, and shared that the ranking system orders results by total
hit count regardless of location, forcing examiners to open documents and
manually scan for relevant sections. P02 used a similar process of scrolling
and combing through CTRL+F results of keywords to find relevant paragraphs.
P01 expressed further frustration with granted patents, which require citing
by column and line number rather than paragraph number, adding manual overhead
when recording where evidence appeared. P09 described that some internal tools
support multi-highlighting to track keyword locations across claims, but
public tools lack this capability entirely, forcing examiners to flip between
separate documents to locate where relevant content appears manually. In
practice, P10 also described relying on scanning titles, images, and
highlighted keywords to triage relevance, further illustrating that examiners
must manually piece together signals across document representations to
determine whether a result is worth deeper inspection. Several participants
also described needing to read beyond the keyword hit itself to determine
whether it was actually meaningful in context, often revisiting multiple
sections of the document to interpret how the term was being used. For
example, P03 explained that when a result appears potentially relevant, they
cannot rely on keyword matches alone and instead ``have to go through the
whole patent'' to verify whether it actually aligns with the claim, since
surface-level indicators like keyword proximity are often insufficient.
Together, these patterns suggest that relevance assessment is slowed not only
by the difficulty of locating keyword matches but also by the additional
cognitive effort required to interpret their significance in complex technical
documents.

\smallskip\noindent\textbf{Need 5: Existing AI-assisted search tools do not
align with the iterative, refinement-based search strategy that examiners
actually use.}
Participants with experience using AI or semantic search tools---including the
USPTO's similarity search, IP.com, and other platforms---consistently
described them as inadequate. P02 emphasized that effective search ``really
involves appraising the art as you search for it, and refining your search
based on that,'' which current tools do not support. P06 described the
similarity search as ``hit or miss,'' identifying the core problem as the
tool's inability to distinguish domain context---a medical-device search for
``user interface'' returns results from banking and unrelated software. P07,
while noting IP.com ``seems to interpret language a little bit more
intuitively,'' still gravitated toward non-patent sources like technical
manuals as a more efficient strategy. P09 echoed the mixed assessment, noting
that the similarity search ``has mixed results---sometimes I just click that
button\ldots\ and it finds me 300 references, of which 2 or 3 are actually
useful. Other times, it's just completely off the mark.'' P09 also described
a workaround of feeding publicly available PG pubs into external AI tools to
help devise search strategies, while noting that proprietary restrictions
prevent using AI with actual application information. It appeared that, across
most participants, AI tools were supplements rather than primary methods,
reinforcing that future design should augment iterative workflows rather than
replace them.

\subsection{Implications for Future Research and Tool Design}

Several of our findings align with patterns reported in prior work on
professional information work, including the challenges of maintaining context
across documents~\cite{Ko2006}, organizing scattered evidence~\cite{Russell1993},
and iterative refinement in expert search~\cite{Pirolli2005}. However, our
findings also diverge from prior work in important ways. First, while Ko
et al.~\cite{Ko2006} found that developers lose track of reasoning chains
across files, patent examiners face an additional structural constraint: their
reasoning must be organized element-by-element against legally defined claim
language, not simply against a mental model of the codebase. The failure mode
is therefore not just lost context but legally indefensible analysis. Second,
Lupu et al.~\cite{Lupu2013} identified relevance assessment as an open problem
in patent retrieval, but framed it primarily as a ranking and interface
problem; our findings show that the deeper issue is the absence of integrated
support for the interpretive work that follows retrieval---claim mapping,
synonym management, and filing-date verification---which existing tools do not
address at all. Third, Storey et al.~\cite{Storey1999} observed that developers
alternate among cognitive strategies and argued that tools should accommodate
this flexibility; we find a similar pattern among examiners, but with the
added constraint that the output of their sensemaking must conform to a rigid
legal structure, limiting how flexibly they can reorganize their analysis.
Finally, while Pirolli and Card's sensemaking model~\cite{Pirolli2005}
describes foraging and sensemaking as iteratively coupled, our participants
described AI-assisted search tools that treat these as separate, one-shot
operations (Need~5), suggesting that current tool design actively works
against the iterative workflow that prior theory predicts experts will use.

These differences suggest that future research should more explicitly account
for domains where information work is not only interpretive but also
constrained. Our findings also suggest that future tool design could
prioritize integration over automation. Rather than replacing examiner
judgment with AI-driven search, the most impactful opportunities may lie in
reducing manual overhead for tasks examiners already perform, such as synonym
management, date filtering, and claim-to-evidence mapping.


%% ====================================================================
%%  SECTION 5 — DESIGN OF THE TOOL
%% ====================================================================
\section{Design of the Tool}

\subsection{Overview}

\smallskip\noindent\textbf{What the tool is.}
Our system is a programmatic document analysis environment in which patent
documents are treated as structured, queryable data rather than opaque text
blobs. An examiner begins by uploading one or more patent documents (PDF or
plain text), which the system parses into a \texttt{Document~$\to$~Section~$\to$~Passage}
hierarchy. The examiner then selects a subset of loaded documents to search
over and writes a query in a lightweight textual DSL. The system executes the
query at the passage level and returns a list of matched passages, each
displayed in a result card. Each card shows: the matched passage text with
the queried terms highlighted; the document title and section type
(\textsc{specification}, \textsc{claims}, etc.); the paragraph or claim
anchor (\texttt{[0042]}, \emph{Claim~3}) if present; the immediately
preceding and following passages for context; and a human-readable
explanation of which predicates matched and why.

The interface is a split-panel layout: a document manager and query editor
on the left, a scrollable list of result cards on the right. Results update
live as the examiner types, with a 600ms debounce, so the result set
reflects the current query before the examiner has finished formulating the
next refinement step. There is no ``run search'' button to press; the tight
feedback loop is intentional.

\smallskip\noindent\textbf{Why this tool, for this population.}
Every design decision in this section is traceable to a specific observation
from our contextual inquiry sessions. The document model reflects the legal
structure of patents as participants described it. The DSL operators map
directly to the reasoning strategies participants actually used. The live
refresh matches the iterative appraise-and-refine cycle that P02 articulated
as the correct mental model for patent search. We describe these connections
in detail below.


\subsection{Walkthrough: P03's Claim-Mapping Task}

To make the design concrete, we walk through how our tool would support a
task we observed during session P03. P03 was working a software patent
application whose independent claim recited a system for \emph{dynamic
routing table updates}. Having identified seven candidate prior art references
via a USPC/CPC class search, P03 needed to determine whether any of them
disclosed the claimed routing mechanism. The process we observed occupied
more than 20 minutes and involved the following manual steps: (1) opening
each reference in a USPTO viewer; (2) using CTRL+F to locate the term
``routing'' anywhere in the document; (3) scrolling through every match
regardless of section---abstract, background, claims---to assess whether it
described a dynamic update mechanism; (4) if promising, manually recording
the paragraph number and section for later citation; and (5) repeating for
every synonym that P03 could think of (``forwarding table,'' ``FIB,''
``routing cache''). P03 noted that of the seven initial results, only one
had a usable filing date, but determining this also required a separate
manual inspection of each.

With our tool, the same task would proceed as follows.

\smallskip\noindent\emph{Step 1: Load and scope the corpus.}
The examiner uploads the seven candidate PDFs. The document manager lists
them with their parsed titles and CPC codes. To immediately filter to a
legally admissible subset, the examiner can write:

\begin{lstlisting}[caption={Filter by filing date and CPC before opening any document.}]
meta.filingDate:<2018-03-15 AND cpc:"H04L 45/00"
\end{lstlisting}

\noindent This eliminates references that postdate the application's
effective filing date without the examiner having to open each document.
This directly addresses Need~2: P03 described having to manually inspect
each of the seven results to determine which had a usable date.

\smallskip\noindent\emph{Step 2: Query across the corpus for the claim element.}
The examiner then writes a query targeting the specific claim language.
Rather than typing one term, inspecting results, and then starting over with
a synonym, the examiner composes the full synonym set in a single query that
executes across all selected documents simultaneously:

\begin{lstlisting}[caption={Synonym expansion across a loaded corpus (Need 1).}]
section:SPECIFICATION AND
  (contains:"routing table" OR contains:"forwarding table"
   OR contains:"FIB" OR contains:"routing cache")
\end{lstlisting}

\noindent The result set now contains only passages from the detailed
specification---the section legally relevant for mapping a claim element to
evidence---that mention any of the four synonymous terms. Passages from the
abstract, background, and claims are excluded. Passages from all selected
documents appear in one scrollable list. This addresses both Need~1 (synonym
management) and Need~4 (section-scoped keyword localization) simultaneously.

\smallskip\noindent\emph{Step 3: Assess results in context.}
Each result card displays the matched passage text alongside the passage
immediately before and after it. P03 observed that keyword hits alone were
often not sufficient to determine relevance---``you have to go through the
whole patent'' to understand whether a term was being used in the right
sense. The neighboring passages eliminate most of that reading burden. For
each card, the paragraph anchor (\texttt{[0042]}) is displayed, ready to be
copied into the claim chart without further document lookup. This addresses
the citation overhead P01 described: the overhead of manually recording
``column and line number'' for granted patents is eliminated because the
anchor is extracted at parse time.

\smallskip\noindent\emph{Step 4: Iterative refinement.}
If the initial results contain false positives (e.g., passages about static
routing that happen to mention ``routing table''), the examiner narrows the
query by adding a term without restarting:

\begin{lstlisting}[caption={Narrowing by adding a required term.}]
section:SPECIFICATION AND
  (contains:"routing table" OR contains:"forwarding table")
  AND contains:"dynamic"
\end{lstlisting}

\noindent The result set updates within 600ms. The examiner can see
immediately whether the added constraint reduced the result set to something
manageable or eliminated too many relevant passages, and can adjust
accordingly. This tight iteration loop was absent from every tool P03,
P05, and P09 described. P02 articulated the correct model: effective search
``really involves appraising the art as you search for it, and refining
your search based on that'' (Need~5). Our live refresh is a direct
implementation of that observation.


\subsection{The Document Abstraction}

\smallskip\noindent\textbf{The hierarchy.}
Each patent document is parsed into a three-level hierarchy:
\texttt{Document~$\to$~Section~$\to$~Passage}. A \texttt{Document} carries
all extracted metadata: title, document identifier, filing and publication
dates, application number, inventors, assignee, CPC classification codes,
and domestic priority chain. A \texttt{Section} has a canonical type drawn
from the set \{\textsc{title}, \textsc{abstract}, \textsc{background},
\textsc{summary}, \textsc{specification}, \textsc{claims}, \textsc{other}\}
and a human-readable heading string extracted from the document. A
\texttt{Passage} is an atomic unit of text within a section: a
blank-line-separated paragraph, a USPTO-numbered paragraph (\texttt{[0042]~...}),
or an individual claim.

This hierarchy was not chosen arbitrarily. During coding we noted that
participants never described their information needs in terms of the full
document; they consistently described them at the level of \emph{sections}
(``I want hits in the spec, not the abstract'') and \emph{passages} (``I
need to know which paragraph to cite''). The three-level model is a direct
structural translation of how patent examiners reason about documents.

\smallskip\noindent\textbf{Why section types are first-class.}
Patent sections are not stylistic divisions; they carry distinct legal
functions that every participant in our study treated as categorical. The
\textsc{abstract} is explicitly excluded as prior art under 35 U.S.C.
\S 102(b)(2)(B) in certain conditions; the \textsc{background} describes the
problem the invention addresses; the \textsc{specification} (detailed
description) is the primary locus of technical teaching and is what examiners
cite when writing a rejection; the \textsc{claims} define the legally
protected scope. P05 was explicit: they wanted hits ``in the background and
description sections, not in the abstract or claims.'' No current tool
surfaces results at this level of section specificity. Our model makes
section type queryable as a first-class predicate.

\smallskip\noindent\textbf{Passage-level anchors.}
Each \texttt{Passage} optionally carries one of three anchor types. A
\texttt{paragraphId} records the USPTO paragraph number (e.g.,\ \texttt{0042})
for numbered paragraphs in post-2001 publications. A \texttt{claimNo}
records the claim number for passages parsed from the claims section. A
\texttt{figureRefs} list records any figure references that appear within
the passage text. The \texttt{paragraphId} directly addresses the citation
overhead P01 described: when writing an office action, examiners must record
exactly where in a reference the evidence appears. For published applications,
this means paragraph numbers; for granted patents, column and line numbers.
P01 called this process manual and frustrating. By extracting and storing
the anchor at parse time, our system can surface it on the result card
without requiring the examiner to open the document.


\subsection{The Query DSL}

The query language is a minimal composable DSL. The full grammar supports
phrase containment, section type, paragraph anchor, CPC code, arbitrary
metadata fields, and Boolean composition (\texttt{AND}, \texttt{OR},
\texttt{NOT}) with parenthesized sub-expressions and standard Boolean
precedence. Every predicate in the language was motivated by a specific
observed need; we describe each with its motivation below.

\smallskip\noindent\textbf{Content filtering (\texttt{contains:}).}
The simplest query is a phrase containment check over all passages in all
selected documents:

\begin{lstlisting}[caption={Basic phrase search.}]
contains:"prior art"
\end{lstlisting}

\noindent This is the direct analog of a keyword search in any existing tool,
but with a critical difference: the result is a list of individual passages,
not a ranked list of documents. The examiner sees where in the document the
term appears without having to open anything.

\smallskip\noindent\textbf{Section filtering (\texttt{section:}).}
Adding a \texttt{section:} predicate restricts results to passages from a
specific document section. This was the single most consistently requested
capability across our study:

\begin{lstlisting}[caption={Section-scoped containment query (Need 4).}]
section:SPECIFICATION AND contains:"virtual machine migration"
\end{lstlisting}

\noindent The predicate accepts any canonical section type and is
case-insensitive. P05 wanted hits ``in the background and description
sections, not in the abstract or claims,'' and described the absence of this
feature as something that ``makes no sense for what I actually need to find.''
This predicate is the direct implementation of that need.

\smallskip\noindent\textbf{Synonym expansion (\texttt{OR}).}
Need~1 established that examiners must manually enumerate synonyms and
run separate queries for each variant. The \texttt{OR} operator allows the
full synonym set to be expressed in a single query:

\begin{lstlisting}[caption={OR for synonym expansion across a corpus (Need 1).}]
section:CLAIMS AND
  (contains:"virtual machine" OR contains:"hypervisor"
   OR contains:"guest operating system")
\end{lstlisting}

\noindent This returns all claim passages from all selected documents that
mention any of the three terms. P04 described maintaining a shared synonym
crosswalk; this query structure makes that crosswalk executable rather than
just a reference document, and applies it across a corpus in one step.

\smallskip\noindent\textbf{Negation (\texttt{NOT}).}
The \texttt{NOT} operator excludes passages from the result set. A practical
use case is filtering the introductory boilerplate paragraph that many
specifications begin with, which often contains broad terminology that
matches almost any query:

\begin{lstlisting}[caption={NOT to exclude boilerplate preamble.}]
section:SPECIFICATION AND contains:"network interface"
  AND NOT paragraph:1
\end{lstlisting}

\smallskip\noindent\textbf{Paragraph anchoring (\texttt{paragraph:}).}
The \texttt{paragraph:} predicate retrieves a specific numbered passage by
its USPTO paragraph anchor. This is useful when an examiner returns to a
document in a subsequent session and needs to pull up a known-relevant
paragraph without re-reading the specification:

\begin{lstlisting}[caption={Pinpoint retrieval by paragraph anchor (Need 4).}]
section:SPECIFICATION AND paragraph:42
\end{lstlisting}

\noindent This also enables cross-document comparison: an examiner who has
noted paragraph \texttt{[0042]} as relevant in one reference can query the
same anchor across other loaded documents to see whether they share
structurally similar placement for the same concept.

\smallskip\noindent\textbf{Metadata filtering (\texttt{meta.} and \texttt{cpc:}).}
CPC classification codes and arbitrary metadata fields narrow the document
scope before content is examined. This reflects the corpus-scoping workflow
that multiple participants described: they first constrain the set of
candidate documents by classification, then search their content.

\begin{lstlisting}[caption={CPC scope then section filter.}]
cpc:"H04L 45/00" AND section:SPECIFICATION
  AND contains:"link-state advertisement"
\end{lstlisting}

\begin{lstlisting}[caption={Date constraint to enforce admissibility (Need 2).}]
meta.filingDate:<2019-06-01
  AND section:CLAIMS AND contains:"neural interface"
\end{lstlisting}

\noindent The date predicate directly addresses Need~2. P05 described
completing an entire office action on a reference that postdated the
application's effective filing date, requiring a complete redo. P06 called
the absence of reliable date filtering ``definitely a very good fix that can
be done.''

\smallskip\noindent\textbf{Full composability.}
All predicates compose with arbitrary nesting. The following query reflects
the kind of compound search P03 assembled across multiple tool sessions
during our study---expressed here in a single executable string that can be
saved, shared, and rerun:

\begin{lstlisting}[caption={Compound query for a single claim element.}]
meta.filingDate:<2018-03-15 AND section:SPECIFICATION AND
  (contains:"routing table" OR contains:"forwarding table"
   OR contains:"FIB")
  AND contains:"dynamic"
\end{lstlisting}

\noindent Each result card shows the matched passage text, the section type
and heading, the paragraph or claim anchor if present, the immediately
preceding and following passages for context, and a match-reason string that
names each predicate that matched (e.g., \texttt{section:SPECIFICATION},
\texttt{contains:"routing table"}). The examiner can use this string directly
as a citation trace in a claim chart, addressing the absence of integrated
claim-mapping support identified in Need~3.


\subsection{Key Design Decisions}

\smallskip\noindent\textbf{Decision 1: Passage granularity with neighboring context.}
The most fundamental question in designing a retrieval tool is the unit of
retrieval. Existing tools return documents; sentence-level tools exist in
academic NLP but are unfamiliar to practitioners. We chose the passage
because it corresponds to how patent documents are legally structured and
how examiners described their information needs: they wanted ``where in the
patent'' a term appeared, not just ``which patent.''

However, our analysis sessions surfaced a secondary problem: a passage
clipped from its context can be misleading or uninterpretable. P03 was
explicit that ``you have to go through the whole patent'' in part because the
keyword match by itself does not reveal whether the surrounding text confirms
or contradicts the relevance judgment. We resolved this by including the
immediately preceding and following passages in every result card. This adds
enough local context for a relevance judgment without requiring the examiner
to load the full document. The tradeoff---longer result cards, more
scrolling---was acceptable because it directly replaces the longer task of
opening each document to read around each hit.

\smallskip\noindent\textbf{Decision 2: Section type as a predicate with a
closed vocabulary, not a metadata field.}
An early design used a generic \texttt{meta.} namespace for all document
properties, which would have encoded section type as
\texttt{meta.sectionType:"SPECIFICATION"}. We changed this to a dedicated
\texttt{section:} predicate with a closed vocabulary after analyzing how
participants talked about sections during the CI sessions.

Participants never referred to sections by arbitrary heading text; they
referred to them by legal role. P05 said ``the background and description
sections,'' not ``the section whose heading contains 'Detailed Description.'''
This suggested that the model needed a closed, normalized vocabulary of legal
section types rather than a string-match against headings---particularly
because heading text varies across issuers and publication formats (e.g.,
``Detailed Description of the Preferred Embodiments'' vs. ``Detailed
Description''). Normalizing to canonical types at parse time and exposing
them through a dedicated predicate syntax reflects the conceptual model
participants actually held.

\smallskip\noindent\textbf{Decision 3: Boolean composition over ranked retrieval.}
We made an early decision to build a Boolean filter model rather than a
ranked retrieval model (e.g., TF-IDF, BM25, or embedding similarity). This
was one of the most consequential choices in the design and was almost
entirely driven by Need~5.

P02 described effective search as ``appraising the art as you search for it,
and refining your search based on that.'' This articulates a specific
epistemic structure: the examiner's understanding of what they are looking for
evolves as they look. Ranked retrieval systems assume the query is fixed and
the model's job is to order the corpus against it; they do not expose
\emph{why} a document ranked where it did, making it opaque which dimension
to refine when results are wrong. P09 described the USPTO similarity search
as producing 300 references ``of which 2 or 3 are actually useful,'' with no
clear signal about what to adjust. P06 identified the core problem: the
model has no awareness of domain context, so a search for ``user interface''
in a medical device application returns banking software results.

A Boolean DSL is transparent by design. Every predicate either matches or
does not, and the match-reason string on each result card names exactly which
predicates fired. When a result is missing, the examiner can add an
\texttt{OR} clause; when there are too many false positives, they can add an
\texttt{AND} constraint. Each edit is a one-token change with an immediately
visible effect. This transparency is precisely what makes it compatible with
the iterative appraise-and-refine strategy that participants described as
their actual method.

\smallskip\noindent\textbf{Decision 4: Textual DSL over a GUI filter panel.}
We considered implementing the same predicate semantics as a graphical filter
panel with dropdowns, checkboxes, and date pickers. The textual DSL won out
for reasons that came directly from the CI sessions.

First, patent examiners already write Boolean query strings as a core job
skill. P04, P07, and P10 each described constructing Boolean queries as
standard practice. A textual DSL is a familiar interaction mode for this
population; a filter panel would introduce a new interaction paradigm for
examiners who already think in terms of query strings.

Second, and more importantly, a textual query is a first-class persistent
artifact in a way that a GUI selection state is not. P01 maintained a Word
table for claim mapping; P03 used inline bullet points; P05 copied claims
into a Word document. All of these are workarounds for the absence of
integrated documentation support (Need~3). A query string can be pasted
directly into a claim chart as a citation trace: ``found via
\texttt{section:SPECIFICATION AND contains:"routing table"}, paragraph
[0042].'' A GUI filter panel cannot be recorded that way. The DSL makes
queries \emph{shareable}---with colleagues, across cases, and over time---in
a way that a stateful UI control does not.

Third, the live debounce refresh means that typing in the text box already
behaves like a responsive filter panel: each character edit produces updated
results within 600ms. The perceived interactivity difference between a
textual DSL and a GUI panel is therefore small, while the artifact value of
the textual form is substantially higher.


%% ====================================================================
%%  SECTION 6 — TECHNICAL CHALLENGES AND SOLUTIONS
%% ====================================================================
\section{Technical Challenges and Technical Solutions}

% TODO: This section should be 2-3 pages. Focus on 2-3 hardest problems
% that required inventing something new---not debugging challenges.
% Frame each challenge formally, then describe the solution.

\subsection{Challenge 1: Parsing Heterogeneous Patent Formats into a Uniform Hierarchy}

% TODO: State the problem formally.
%
% Patent documents arrive in two formats: plain-text files (.txt) with
% explicit section headings and bracket-numbered paragraphs, and
% text-extracted PDFs whose formatting is far less predictable. The parser
% must map both into the same Document -> Section -> Passage abstraction
% without per-document manual intervention.
%
% Key sub-problems:
%   - Section heading normalization: "BRIEF DESCRIPTION" vs. "BRIEF SUMMARY"
%     vs. "BACKGROUND OF THE INVENTION" must all map to a canonical type.
%   - Passage boundary detection: paragraph [NNNN] anchors in .txt files
%     vs. inferred paragraph breaks in PDFs.
%   - Anchor extraction: paragraphId, claimNo, figureRefs are structurally
%     present in .txt but must be heuristically inferred from PDF text streams.
%
% Describe the solution: the heading normalization table, the state-machine
% parser for section transitions, and the anchor annotation pass.
% Quantify: how many heading variants were encountered across the 5 PDFs
% pulled in from data/raw/? What was the fallback strategy?

\subsection{Challenge 2: Designing a Composable Filter Model for the DSL}

% TODO: State the problem formally.
%
% The DSL must support arbitrarily nested AND/OR combinations over
% heterogeneous predicates (section type, text containment, CPC code,
% metadata fields, paragraph anchor). The challenge is to define a
% compositional semantics for the filter language such that:
%   (a) every well-formed query has a well-defined denotation over the
%       passage set of a document, and
%   (b) the execution engine can evaluate queries efficiently without
%       materializing the full cross-product of (document, passage) pairs.
%
% Describe the solution: the internal query IR (the Filter ADT / query tree),
% how the DSL parser maps surface syntax to the IR, and how the execution
% engine evaluates the tree depth-first over the passage stream.
% Include a small diagram or pseudocode of the IR node types if helpful.

\subsection{Challenge 3: Provenance and Context Reconstruction}

% TODO: State the problem formally.
%
% A matched passage is only useful to an examiner if they can quickly
% understand (a) where in the document it appears, (b) what surrounds it,
% and (c) why it matched. This requires that results carry provenance
% metadata (section type, paragraphId/claimNo, document id, passage index)
% and that neighboring passages can be reconstructed without re-parsing the
% document.
%
% Describe the solution: how the parsed JSON document store preserves index
% offsets and anchor annotations, and how the query engine attaches
% neighboring passages to each match result.

\subsection{Remaining Work}

% TODO: List what remains to be done in the prototype:
%   - Filing-date filtering (meta.filing_date predicate) with proper
%     priority-date resolution (Need 2)
%   - Synonym expansion operator or integration with a thesaurus backend
%     (Need 1)
%   - Claim-element tagging and mapping workspace (Need 3)
%   - Support for NOT and parenthesized precedence in the DSL grammar
%   - Embedding-based retrieval as an alternative ranking mode


%% ====================================================================
%%  SECTION 7 — IMPLEMENTATION
%% ====================================================================
\section{Implementation}

% TODO: 1 paragraph to 0.5 pages of low-level implementation details.
%
% Suggested content:
%   - Backend: Python 3, FastAPI, pypdf for text extraction from PDFs,
%     custom recursive-descent DSL parser, passage store in structured JSON.
%   - Frontend: React + TypeScript, Vite, thin client that calls
%     /documents, /documents/parse, and /query endpoints.
%   - Parser pipeline: parse_raw_to_json script ingests from data/raw/,
%     writes structured JSON to data/parsed/.
%   - Repository: https://github.com/shanayamalik/cs294-sp26
%   - Tests: pytest unit tests for the DSL parser, query engine, and
%     source loader.
%   - Supported document sources: .txt and text-extractable .pdf;
%     image-only/scanned PDFs require OCR pre-processing.

The backend is implemented in Python~3 using FastAPI and exposes three
endpoints: \texttt{/documents} (list loaded documents), \texttt{/documents/parse}
(trigger parsing of raw files), and \texttt{/query} (execute a DSL query).
The frontend is a thin React/TypeScript client built with Vite. The parser
pipeline uses \texttt{pypdf} for text extraction from PDFs and a custom
recursive-descent parser for the DSL. Structured JSON documents are stored
in \texttt{backend/data/parsed/}. The full implementation is available at
\url{https://github.com/shanayamalik/cs294-sp26}.

% TODO: Add sentences on test coverage, known limitations (OCR not supported),
% and approximate lines of code or module count if useful.


%% ====================================================================
%%  SECTION 8 — DISCUSSION
%% ====================================================================
\section{Discussion}

% TODO: ~1 page. Answer:
%   - What are the key takeaway lessons from building this tool?
%   - How could future implementers of languages or tools use these lessons?
%
% Suggested themes:
%
%   1. Structuring document formats as typed hierarchies enables a class of
%      queries that are impossible in flat-text search systems. The legal
%      structure of patents (sections with fixed roles, numbered claims,
%      paragraph anchors) is a resource that existing tools ignore but that
%      our tool exploits.
%
%   2. The gap between retrieval and sensemaking is a design space, not just
%      a product gap. Building a DSL that lives in this gap (post-retrieval,
%      pre-mapping) reveals that the two stages are not naturally sequential
%      but iteratively coupled---a lesson for future tool builders in any
%      domain with similar structure (legal documents, scientific literature,
%      regulatory filings).
%
%   3. User-centered design at the DSL level: the need-finding work directly
%      shaped the choice of section-scoped predicates, the AND/OR composability
%      (vs. ranked retrieval), and the passage-level granularity of results.
%      Without the study, it would have been tempting to build a simpler
%      full-text search interface.


%% ====================================================================
%%  SECTION 9 — WHAT ELSE SHOULD YOU KNOW?
%% ====================================================================
\section{Additional Notes}

% TODO: 1-3 pages of free-form content. Suggested topics:
%
%   (a) Explicitly trace the connection from each of the five needs to
%       specific tool features, for any reader who might still be uncertain
%       about the HCI-PL connection:
%
%       Need 1 (synonyms)    -> OR operator in DSL for synonym expansion
%       Need 2 (filing date) -> meta.filing_date predicate (planned)
%       Need 3 (claim mapping) -> passage-level results with claimNo anchors,
%                                  query reuse as text artifacts
%       Need 4 (keyword location) -> section-scoped queries, neighboring context
%       Need 5 (iterative refinement) -> composable DSL allows incremental
%                                         query construction and reuse
%
%   (b) Limitations and threats to validity:
%       - The prototype has been evaluated through design review, not a
%         follow-up user study; it is not yet validated with patent examiners.
%       - PDF parsing quality varies with document layout.
%       - The document store currently holds a small corpus (5 patents);
%         scalability has not been tested.
%
%   (c) Broader applicability:
%       - The Document -> Section -> Passage abstraction and the DSL design
%         generalize beyond patents to other structured document corpora
%         (regulatory filings, scientific papers, legal opinions).
%       - The approach of exposing document structure as typed filter
%         predicates could inform DSL design in adjacent domains.
%
%   (d) Future directions:
%       - Embedding-based retrieval as a soft-match alternative to exact
%         contains: predicates (Need 1 / Need 5)
%       - A claim workspace that integrates DSL query results directly into
%         a structured claim chart (Need 3)
%       - USPTO search integration for live prior-art retrieval within the
%         same query environment


\begin{acks}
We thank Professor Sarah E.\ Chasins for her guidance throughout this
project and the CS~294 course at UC Berkeley. We also thank the ten patent
examiners who generously shared their time, expertise, and workflows with us.
\end{acks}

\begin{thebibliography}{12}

\bibitem{Ko2006}
A.~J. Ko, B.~A. Myers, M.~J. Coblenz, and H.~H. Aung.
\newblock An Exploratory Study of How Developers Seek, Relate, and Collect
  Relevant Information during Software Maintenance Tasks.
\newblock {\em IEEE Trans. Softw. Eng.}, 32(12):971--987, 2006.

\bibitem{Pirolli2005}
P.~Pirolli and S.~Card.
\newblock The Sensemaking Process and Leverage Points for Analyst Technology
  as Identified Through Cognitive Task Analysis.
\newblock In {\em Proc. Int. Conf. Intelligence Analysis}, volume~5.
  MITRE, McLean, VA, 2005.

\bibitem{Sadowski2018}
C.~Sadowski, E.~S{\"o}derberg, L.~Church, M.~Sipko, and A.~Bacchelli.
\newblock Modern Code Review: A Case Study at Google.
\newblock In {\em Proc. ICSE-SEIP '18}, pages 181--190. ACM, 2018.

\bibitem{Russell1993}
D.~M. Russell, M.~J. Stefik, P.~Pirolli, and S.~K. Card.
\newblock The Cost Structure of Sensemaking.
\newblock In {\em Proc. CHI '93}, pages 269--276. ACM, 1993.

\bibitem{Lupu2013}
M.~Lupu, K.~Mayer, N.~Kando, and A.~J. Trippe.
\newblock {\em Current Challenges in Patent Information Retrieval},
  volume~29 of {\em The Information Retrieval Series}.
\newblock Springer, 2013.

\bibitem{Storey1999}
M.-A.~D. Storey, F.~D. Fracchia, and H.~A. M{\"u}ller.
\newblock Cognitive Design Elements to Support the Construction of a Mental
  Model during Software Exploration.
\newblock {\em J. Systems and Software}, 44(3):171--185, 1999.

\bibitem{Nersessian2008}
N.~J. Nersessian.
\newblock {\em Creating Scientific Concepts}.
\newblock MIT Press, 2008.

\bibitem{Beyer1997}
H.~Beyer and K.~Holtzblatt.
\newblock {\em Contextual Design: Defining Customer-Centered Systems}.
\newblock Morgan Kaufmann, 1997.

\bibitem{Lubin2021}
J.~Lubin and S.~E. Chasins.
\newblock How Statically-Typed Functional Programmers Write Code.
\newblock {\em Proc. ACM Program. Lang.}, 5(OOPSLA):Article 155, 2021.

\bibitem{Zhang2020}
T.~Zhang, B.~Hartmann, M.~Kim, and E.~Paez.
\newblock Enabling Data-Driven API Design with Community Usage Data:
  A Need-Finding Study.
\newblock In {\em Proc. CHI '20}, pages 1--13. ACM, 2020.

\bibitem{Braun2006}
V.~Braun and V.~Clarke.
\newblock Using Thematic Analysis in Psychology.
\newblock {\em Qualitative Research in Psychology}, 3(2):77--101, 2006.

% TODO: Add citations for new related work covered in Section 2
% (DSL-related work, related tools). Suggested additions:
%
% \bibitem{Elasticsearch2018}
% C.~Gormley and Z.~Tong.
% \newblock {\em Elasticsearch: The Definitive Guide}.
% \newblock O'Reilly Media, 2015.
%
% \bibitem{Chamberlin1974}
% D.~D. Chamberlin and R.~F. Boyce.
% \newblock SEQUEL: A Structured English Query Language.
% \newblock In {\em Proc. ACM SIGFIDET Workshop}, pages 249--264. ACM, 1974.

\end{thebibliography}

\end{document}
